Assessing the Impact of MDE and Code Generation on the Sustainability of Scientific Computing Software: A Research Proposal

### 2019-09-11 ResearchProposal.tex
Additional changes to structure, additions to background, editing of introduction

\usepackage[round]{natbib}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=green,       % color of links to bibliography
    filecolor=magenta,   % color of file links
    urlcolor=cyan           % color of external links
}

@@

@@

Quality of SCS improved through documentation.  SCS developers do not always
value documentation, but grants such as, NSERC's CRDPJ 447533-13, have allowed
me to demonstrate its value and to provide techniques to improve its quality.
For instance, \citet{SmithEtAl2015-SS-TR}
shows that the quality of statistical software for psychology is generally
improved when developed using the structured CRAN (Comprehensive R Archive
Network) process and tools, versus an ad hoc process.
\citet{SmithAndKoothoor2016}
highlights the value of proper documentation by redeveloping nuclear safety
analysis software.  Twenty seven (27) worrisome documentation problems were
found, including incompleteness, ambiguity, inconsistency, verifiability,
modifiability, traceability and a lack of abstraction.  A redevelopment
experiment with five existing projects \citep{SmithJegatheesanAndKelly2016}
enabled the code owners (as ascertained through interviews) to clearly see the
value of documentation.  However, mirroring other studies
\citep{CarverEtAl2007}, the code owners felt documentation takes too much time.

One promising approach to avoid the impending collapse is to use a Model Driven
Engineering (MDE) where MDE is an engineering approach that exploits structured representations
(metamodels for abstract syntax, and constraints to capture static semantics) to
automate repetitive, tedious and error prone tasks.  In particular, code
generation.   approach together with code generation.  Citation supporting
this idea.  My research will dramatically reduce the cost of developing reliable,
reproducible and re-certifiable software by: i) creating infrastructure for a
knowledge base of scientific and computing models; and, ii) writing explicit
``recipes'' that weave together this knowledge to generate theoretical models,
design documents, code, test cases and build scripts. The generator will render
to multiple languages, such as html, LaTeX, RTF for documentation and Python,
Java, C, and Matlab, for code. One source of knowledge, with rules for
transformation, means completeness, consistency and traceability can be achieved by
construction. Moreover, these qualities can be maintained as requirements are
modified, design decisions are changed, documentation standards are varied, and
software is re­certified.  The knowledge will be built up incrementally.  New
projects will reuse existing knowledge and expand as necessary.

@@

\section{Literature Review}

roadmap of this section

@@

- how currently done, qualities of interest, add sustainability

@@

@@

@@

\subsection{Medical Imaging}

Medical imaging and analysis software (Dr.\ Michael Noseworthy).

\subsection{Climate Modelling Software}

Climate modelling software (Dr.\ Zoe Li or Dr.\ Xander Wang).

\subsection{Computational Medicine}

Extraction of 3D geometries from medical images (Dr.\ Zahra Motamed).

\subsection{Finite Element Analysis}

Finite element solvers and/or mesh generators (Dr.\ Dieter Stolle).  Potential
software includes FeNiCS, MOOSE, FreeFEM++, Dolphin, etc.  Install the packages
and use them to do something non-trivial.

\subsection{Psychology Software}

Statistics software for psychology (Dr.\ Karin Humphreys).

\subsection{Chemical Engineering Software}

Li Xi (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi}).

\subsection{Geographic Information Systems}

Geographic Information Systems (Jason Brodeur and/or Patrick DeLuca).

\subsection{Solidification and Casting Software}

Solidification and casting software (Dr.\ Andr\'e Phillion).

@@

Quantum chemistry software (Dr.\ Paul Ayers).

>>>>

\usepackage[round]{natbib}
\usepackage{booktabs}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=blue,       % color of links to bibliography
    filecolor=magenta,   % color of file links
    urlcolor=cyan           % color of external links
}

@@

\newpage

@@

Documentation can address the confidence problem mentioned above.  For software
in general, documentation, written before and during development, provides many
benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
about requirements, more useful design reviews, easier integration of separately
written modules, more effective code inspection, more effective testing, and
more efficient corrections and improvements.  For SCS in particular, developers
do not always value documentation, but documentation has been shown to help
improve SCS.  For instance, \citet{SmithEtAl2015-SS-TR} shows that the quality
of statistical software for psychology is generally improved when developed
using the structured CRAN (Comprehensive R Archive Network) process and tools,
versus an ad hoc process.  \citet{SmithAndKoothoor2016} highlights the value of
proper documentation by redeveloping nuclear safety analysis software.  Twenty
seven (27) worrisome documentation problems were found, including
incompleteness, ambiguity, inconsistency, verifiability, modifiability,
traceability and a lack of abstraction.  A redevelopment experiment with five
existing projects \citep{SmithJegatheesanAndKelly2016} enabled the code owners
(as ascertained through interviews) to clearly see the value of documentation.
However, mirroring other studies \citep{CarverEtAl2007}, the code owners felt
documentation takes too much time.

One promising approach to gain the benefits of documentation, without the often
infeasible time commitment, is to use a Model Driven Engineering (MDE), where
MDE is an engineering approach that exploits structured representations
(metamodels for abstract syntax, and constraints to capture static semantics) to
automate repetitive, tedious and error prone tasks.  In this research proposal,
we will use MDE as a short-hand to include model transformation, Domain Specific
Languages (DSLs) and code generation.  Several current papers point to DSLs and
code/document generation as a transformative technology for documentation,
design and verification of SCS \citep{JohansonAndHasselbring2018, Smith2018}.
The benefit of an MDE approach is to dramatically reduce the cost of developing
reliable, reproducible and re-certifiable software by: i) creating
infrastructure for a knowledge base of scientific and computing models; and, ii)
writing explicit ``recipes'' that weave together this knowledge to generate
theoretical models, design documents, code, test cases and build scripts. The
generator can render to multiple languages, such as html, LaTeX, RTF for
documentation and Python, Java, C, and Matlab, for code. One source of
knowledge, with rules for transformation, means completeness, consistency and
traceability can be achieved by construction. Moreover, these qualities can be
maintained as requirements are modified, design decisions are changed,
documentation standards are varied, and software is re­certified.  The knowledge
will be built up incrementally.  New projects will reuse existing knowledge and
expand as necessary.

@@

\section{Background and Literature Review}

roadmap of this section.  Some overview of the topics themselves, but an
emphasis on how they appear in the SCS context.

\subsection{Software Qualities of Interest}

For instance,
a rational design process can provide quality improvements, and Quality
Assurance (QA), as shown in Table~\ref{Tbl_Qualities}.  Moreover, documentation,
written before and during development, can provide many
benefits~\cite{Parnas2010}: easier reuse of old designs, better communication
about requirements, more useful design reviews, easier integration of separately
written modules, more effective code inspection, more effective testing, and
more efficient corrections and improvements.

- qualities of interest, add sustainability

\begin{description}%[leftmargin=\the\parindent]

\item [\textbf{Verifiability}] involves ``solving the equations
  right''~\cite[p.~23]{Roache1998}; it benefits from rational documentation
  that systematically shows, with explicit traceability, how the governing
  equations are transformed into code.

\item [\textbf{Validatability}] means ``solving the right
  equations''~\cite[p.~23]{Roache1998}.  Validatability is improved by a
  rational process via clear documentation of the theory and assumptions, along
  with an explicit statement of the systematic steps required for experimental
  validation.

% \item [\textbf{Reliability}] is a critical quality for scientific software,
%   since the results of computations are meaningless, if they are not dependable.
%   Reliability is closely tied to verifiability, since the key quality to verify
%   is reliability, while the act of verification itself improves reliability.

% \item [\textbf{Performance}] considerations can make certification challenging, since QA
%   becomes more difficult for more complex code.  However,
%   as Roache~\cite[p.~355]{Roache1998} points out, using simpler algorithms and
%   reducing the number of options in general purpose code, is not always a
%   practical option.

\item [\textbf{Usability}] can be a problem.  Different users, solving the same
  physical problem, using the same software, can come up with different answers,
  due to differences in parameter selection~\cite[p.~370]{Roache1998}.  To
  reduce misuse, a rational process must state expected user characteristics,
  modelling assumptions, definitions and the range of applicability of the code.

\item [\textbf{Maintainability}] is necessary in scientific software, since change,
  through iteration, experimentation and exploration, is inevitable.  Models of
  physical phenomena and numerical techniques necessarily evolve over
  time~\cite{CarverEtAl2007, SegalAndMorris2008}.  Proper documentation,
  designed with change in mind, can greatly assist with change management.%   QA
  % activities need to take the need for creativity into account, while not
  % smothering it~\cite[p.~352]{Roache1998}.

\item [\textbf{Reusability}] provides support for the quality of reliability,
  since reliability is improved by reusing trusted components~\cite{Dubois2005}.
  (Care must still be taken with reusing trusted components, since blind reuse
  in a new context can lead to errors, as dramatically shown in the Ariane 5
  disaster~\cite[p.~37--38]{OliveiraAndStewart2006}.)  The odds of reuse are
  improved when it is considered right from the start.

\item [\textbf{Understandability}] is necessary, since reviewers can only certify
  something they understand.  Scientific software developers have the
  view ``that the science a developer embeds in the code must be apparent to
  another scientist, even ten years later''~\cite{Kelly2013}.
  Understandability applies to the documentation and code, while usability
  refers to the executable software.  Documentation that follows a rational
  process is the easiest to follow.

\item [\textbf{Reproducibility}] is a required component of the scientific
  method~\cite{Davison2012}.  Although QA has, ``a bad name among
  creative scientists and engineers''~\cite[p.~352]{Roache1998}, the community
  need to recognize that participating in QA management also improves
  reproducibility.  Reproducibility, like QA, benefits from a consistent and
  repeatable computing environment, version control and separating code from
  configuration/parameters~\cite{Davison2012}.
\end{description}

\subsection{Desirable Qualities of Documentation}

To address the challenges for adequate documentation for QA and certification
efforts for SCS, the qualities described below need to be a priority.  All but
the final quality listed (abstraction), are adapted from the IEEE recommended
practise for producing good software requirements~\cite{IEEE1998}.

\begin{itemize}

\item \emph{Completeness}: Documentation is said to be complete when all the
  requirements of the software are detailed. That is, each goal, functionality,
  attribute, design constraint, value, data, model, symbol, term (with its unit
  of measurement if applicable), abbreviation, acronym, assumption and
  performance requirement of the software is defined.  The software's response
  to all classes of inputs, both valid and invalid and for both desired and
  undesired events, also needs to be specified.

\item \emph{Consistency}: Documentation is said to be consistent when no subset
  of individual statements are in conflict with each other. That is, a
  specification of an item made at one place in the document should not
  contradict the specification of the same item at another location.

\item \emph{Modifiability}: The documentation should be developed in such a way
  that it is easily modifiable so that likely future changes do not destroy the
  structure of the document. Also it should be easy to reflect the change,
  wherever needed in the document to maintain consistency, traceability and
  completeness. For documentation to be modifiable, its format must be
  structured in a way that repetition is avoided and cross-referencing is
  employed.

\item \emph{Traceability}: Documentation should be traceable, as this
  facilitates maintenance and review. If a change is made to the design or code
  of the software, then all the documentation relating to those segments have to
  be modified.  This property is also important for recertification.

\item \emph{Unambiguity}: Documentation is said to be unambiguous only when
  every requirement's specification has a unique interpretation.  The
  documentation should be unambiguous to all audiences, including developers,
  users and reviewers.

\item \emph{Correctness}: There is no direct tool or method for measuring
  correctness. One way of building confidence in correctness is by reviewing to
  ensure that each requirement stated is one that the stakeholders and experts
  desire.  By maintaining traceability, consistency and unambiguity, we can
  reduce the occurrence of errors and make the goal of reviewing for correctness
  easier.

\item \emph{Verifiability}: Every requirement in the documentation must be the
  one fulfilled by the implemented software. Therefore all the requirements
  should be clear, unambiguous and testable, so that a person or a machine can
  verify whether the software product meets the requirements.

\item \emph{Abstract}: Documented requirements are said to be abstract if they
  state what the software must do and the properties it must possess, but do not
  speak about how these are to be achieved. For example, a requirement can
  specify that an Ordinary Differential Equation (ODE) must be solved, but it
  should not mention that Euler's method should be used to solve the ODE. How to
  accomplish the requirement is a design decision, which is documented during
  the design phase.

\end{itemize}

@@

@@

\subsection{Documentation for SCS}

Table~\ref{TblDocuments} shows the recommended documentation for a scientific
software project.  The documents are typical of what is suggested for scientific
software certification, where certification consists of official recognition by
an authority, or regulatory body, that the software is fit for its intended use.
For instance, the Canadian Standards Association (CSA) requires a similar set of
documents for quality assurance of scientific programs for nuclear power
plants~\cite{CSA1999}.

\begin{table}[!h]
  \caption{Recommended Documentation} \label{TblDocuments}

  \begin{tabular}{l p{8cm}}
    %\hline
    %\toprule
    \textit{Problem Statement} & Description of problem to be solved\\
    \textit{Development Plan} & Overview of development process/infrastructure\\
    \textit{Requirements} & Desired functions and qualities of the software\\
    \textit{V\&V Plan} & Verification that all documentation artifacts,
                           including the code, are internally correct.
                           Validation, from an external viewpoint, that the right
                           problem, or model, is being solved.\\
    \textit{Design Specification} & Documentation of how the requirements are to
                                    be realized, through both a software
                                    architecture and detailed design of modules
                                    and their interfaces\\
    \textit{Code} & Implementation of the design in code\\
    \textit{V\&V Report} & Summary of the V\&V efforts,
                             including testing\\
    \textit{User Manual} & Instructions on installation, usage; worked examples\\
    %\hline
    %\bottomrule
  \end{tabular}
\end{table}

To achieve the qualities listed in Table~\ref{Tbl_Qualities}, the documentation
in Table~\ref{TblDocuments} should have the following qualities:
\emph{complete}, \emph{correct}, \emph{consistent}, \emph{modifiable},
\emph{traceable}, \emph{unambiguous}, and \emph{verifiable}.  All of these
qualities are listed in the IEEE recommended practice for software
requirements~\cite{IEEE1998}.  The IEEE guidelines are for requirements, but
most qualities are relevant for all documentation artifacts.  Another relevant
quality, which is not on the IEEE list, is \emph{abstract}.  Requirements should
state what is to be achieved, but be silent on how it is to be achieved.
Abstraction is an important software development principle for dealing with
complexity~\cite[p.~40]{GhezziEtAl2003}.  Smith and Koothoor present further
details on the qualities of documentation for scientific
software~\cite{SmithAndKoothoor2016}.

@@

Determine criteria for what makes a domain a good fit.

@@

\subsubsection{Medical Imaging}

Medical imaging and analysis software (Dr.\ Michael Noseworthy)?

\subsubsection{Climate Modelling Software}

Climate modelling software (Dr.\ Zoe Li or Dr.\ Xander Wang)?

\subsubsection{Computational Medicine}

Extraction of 3D geometries from medical images (Dr.\ Zahra Motamed)?  Lattice
Boltzmann Solver?

\subsubsection{Finite Element Analysis}

Finite element solvers and/or mesh generators (Dr.\ Dieter Stolle)?  Potential
software includes FeNiCS, MOOSE, FreeFEM++, Dolphin, etc.  Install the packages
and use them to do something non-trivial.

\subsubsection{Psychology Software}

Statistics software for psychology (Dr.\ Karin Humphreys)?

\subsubsection{Chemical Engineering Software}

Li Xi (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi})?

\subsubsection{Geographic Information Systems}

Geographic Information Systems (Jason Brodeur and/or Patrick DeLuca)?

\subsubsection{Physical Metallurgy Software}

Solidification and casting software (Dr.\ Andr\'e Phillion)?

@@

Quantum chemistry software (Dr.\ Paul Ayers)?
